\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-07-02 manuscript_v6/main.tex
updated methods

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, boxrule=0pt, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

Reasons given for this inequality in vaccination uptake have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{17},\protect\hyperlink{ref-mosby2021}{22}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{15}}, and
the influence of misinformation and
disinformation\textsuperscript{\protect\hyperlink{ref-mosby2021}{22}--\protect\hyperlink{ref-freeman2020}{24}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{25}}, lack
of vaccine access in areas where underrepresented groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7}}, as well
as an absence of prioritization of areas inhabited by vulnerable
groups\textsuperscript{\protect\hyperlink{ref-nguyen2021}{26}}. Other
studies have also shown heterogeneity in vaccine uptake within small
governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{27}--\protect\hyperlink{ref-bhuiyan2022}{30}},
indicating that accounting for geographical differences in vaccination
can help predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{31}}.

However, such analyses have been carried out mostly in territories
outside of Canada, where available studies have been focused in certain
cities (such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{32}}, or
Montreal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{33}}), or
have explored differences at a province-wide
level\textsuperscript{\protect\hyperlink{ref-guay2022}{19}}. Therefore,
there is a need for studies that explore spatial differences in
vaccination within the Canadian territory and that consequently, can
help identify disparities that need to be addressed within specific
areas in each province.

This need is especially important in the case of Ontario, the most
populated province of Canada. Between 2007 and 2019, Ontario managed
healthcare access to its inhabitants using 14 intra-provincial divisions
called the Local Health Integration Networks (LHINs), which aimed to
provide an integrated health system for the province. However, this
approach was complex, bureaucratic, and resulted in excessive
expenditures, disparities in mortality rates, the deterioration of
certain performance indicators such as wait times and hospital
readmissions, fragmented electronic health systems, and inequities in
health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{34}--\protect\hyperlink{ref-lysyk2016}{38}}.
Therefore, with the intent of better organizing and delivering care in
late 2019 the provincial government eliminated the LHINs and
incorporated the areas covered by them into six larger Health Regions
(North East, North West, Central, Toronto, West, and East), which are
managed by the new government agency Ontario Health
(OH)\textsuperscript{\protect\hyperlink{ref-dong2022}{36}}.

On the other hand, public health in Ontario is administered by Public
Health Ontario (PHO), a government agency established in late 2007 and
currently composed of 34 Public Health Units (PHUs) that cover the
entire geography of the province. Although these PHUS were commissioned
with leading the distribution of the COVID-19 vaccine within the areas
they covered\textsuperscript{\protect\hyperlink{ref-rotenberg2021}{39}},
there was significant overlap between PHO and OH in the actual vaccine
rollout, and in many instances vaccines were administered in vaccination
clinics ran by members of OH working alone or in conjunction with
members of
PHO\textsuperscript{\protect\hyperlink{ref-ashcroft2023}{40}}.

\begin{itemize}
\tightlist
\item
  add about intersection of Health Regions and PHUs. How vaccination
  presented a challenge as vaccine was ultimately administered in
  facilites managed by OHT?
\end{itemize}

https://www.mcmasterforum.org/docs/default-source/product-documents/rapid-responses/examining-the-intersections-between-ontario-health-teams-and-public-health.pdf?sfvrsn=69caa8c9\_11

Because of the relatively recent adoption of the Health Region model and
its alignment with the onset of the COVID-19 pandemic, there is a need
to analyze if there are ongoing disparities in health access under this
approach that need to be addressed before they are exploited by a new
disease or public health threat. In this regard, previous research has
highlighted disparities in the level of activity of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{41}}.
Therefore, analyzing differences in vaccination uptake within the Health
Regions and can help identify which socio-demographic groups are the
most vulnerable and what areas of the province deserve special attention
by decision-makers.

Therefore, in this study we aim to understand the differences in
vaccination uptake between the different Health Regions of Ontario
between October of 2021 and January of 2022. By including socio-economic
factors in our analysis, we aimed to identify in which groups these
differences were significant in order to provide an assessment of the
current state of healthcare access in Ontario.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto (under protocol 00043317), and which ran between September 30th,
2021 and January 17th,2022. The survey collected socio-economic
information from participants (Table~\ref{tbl-descriptive-stats}), their
location (nearest municipality, as shown in Figure~\ref{fig-map}), the
date of access to the survey, and asked information on vaccination
status by using the question ``Have you received the first dose of the
COVID vaccine?'', with possible answers ``yes'' and ``no''. The original
dataset contained 39,029 observations.

By design, the survey allowed respondents to exit at any time and
deployed the questions randomly. This resulted in \(\approx\) 84\% of
the observations having multiple missing answers or being incomplete due
to participants leaving the survey at different stages. Therefore, we
selected 6,343 observations that were labelled as ``complete'' in the
dataset and that had answers for all covariates considered in our
analysis. Later, we matched the city of each observation with its
corresponding LHIN and Health Region, and removed observations from
areas with low representation (254 observations corresponding to the
North West and North East Health Regions). After all the preliminary
analyses, the total number of observations used for analysis was 6,236
and included the East, Central, Toronto, and West Health Regions
covering between October 1st,2021 and January 17, 2022. The original
dataset, clean dataset, and details on the data cleaning process are
described in detail in the
\href{https://github.com/aimundo/Fields_COVID-19/}{GitHub repository}
for this paper.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{42}--\protect\hyperlink{ref-cnat2022a}{44}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{45}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{46}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{47}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{48}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{49}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{50}}.

@@

Our results show that indeed, there were differences in vaccination odds
across Ontario in certain socio-demographic groups. Specifically, those
who identified as Arab/Middle Eastern, Black, Latin American, having
mixed racial or ethnic background, or that belonged to other groups not
explicitly included in the survey (Southeast Asian, Filipino, West
Asian, and minority groups not identified elsewhere) had vaccination
odds that were between a third and a half of that of individuals that
identified as White or Caucasian (Figure~\ref{fig-models}). These
results are consistent with previous studies that have shown lower
vaccination rates in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{51}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{44},\protect\hyperlink{ref-basta2022}{52}--\protect\hyperlink{ref-cnat2023}{54}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{55}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{56}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{57}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{58},\protect\hyperlink{ref-ct2021}{59}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{60}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{57}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{61}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{62}--\protect\hyperlink{ref-timony2022}{64}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{65}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{66},\protect\hyperlink{ref-eissa2021}{67}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{68}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{65}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{69}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{70}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tsasis2012}{}}%
\CSLLeftMargin{34. }%
\CSLRightInline{Tsasis P, Evans JM, Owen S. Reframing the challenges to
integrated care: A complex-adaptive systems perspective.
\emph{International Journal of Integrated Care}. 2012;12(5).
doi:\href{https://doi.org/10.5334/ijic.843}{10.5334/ijic.843}}

\leavevmode\vadjust pre{\hypertarget{ref-muratov2018}{}}%
\CSLLeftMargin{35. }%
\CSLRightInline{Muratov S, Lee J, Holbrook A, et al. Regional variation
in healthcare spending and mortality among senior high-cost healthcare
users in ontario, canada: A retrospective matched cohort study.
\emph{{BMC} Geriatrics}. 2018;18(1).
doi:\href{https://doi.org/10.1186/s12877-018-0952-7}{10.1186/s12877-018-0952-7}}

\leavevmode\vadjust pre{\hypertarget{ref-dong2022}{}}%
\CSLLeftMargin{36. }%
\CSLRightInline{Dong L, Sahu R, Black R. Governance in the
transformational journey toward integrated healthcare: The case of
ontario. \emph{Journal of Information Technology Teaching Cases}.
Published online December 2022:204388692211473.
doi:\href{https://doi.org/10.1177/20438869221147313}{10.1177/20438869221147313}}

\leavevmode\vadjust pre{\hypertarget{ref-lysyk2015}{}}%
\CSLLeftMargin{37. }%
\CSLRightInline{Auditor General of Ontario O of the, ed. Annual {R}eport
2015. In: \emph{Section 3.08: LHINs - Local Health Integration
Networks}. Queen's Printer for Ontario; 2015. Accessed May 12, 2023.
\url{https://www.auditor.on.ca/en/content/annualreports/arreports/en15/3.08en15.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-lysyk2016}{}}%
\CSLLeftMargin{38. }%
\CSLRightInline{Auditor General of Ontario O of the, ed. Annual {R}eport
2016. In: \emph{Section 3.03: Electronic Health Records' Implementation
Status}. Queen's Printer for Ontario; 2016. Accessed May 12, 2023.
\url{https://www.auditor.on.ca/en/content/annualreports/arreports/en16/v1_303en16.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-rotenberg2021}{}}%
\CSLLeftMargin{39. }%
\CSLRightInline{Rotenberg S, Downer MB, Brown H, et al. \emph{{COVID}-19
Vaccination for People with Disabilities}. Ontario {COVID}-19 Science
Advisory Table; 2021.
doi:\href{https://doi.org/10.47326/ocsat.2021.02.35.1.0}{10.47326/ocsat.2021.02.35.1.0}}

\leavevmode\vadjust pre{\hypertarget{ref-ashcroft2023}{}}%
\CSLLeftMargin{40. }%
\CSLRightInline{Ashcroft R, Donnelly C, Lam S, et al. A qualitative
examination of primary care team's participation in the distribution of
the {COVID}-19 vaccination. Published online June 2023.
doi:\href{https://doi.org/10.21203/rs.3.rs-3016276/v1}{10.21203/rs.3.rs-3016276/v1}}

>>>>

\date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, boxrule=0pt, sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

Reasons given for this inequality in vaccination uptake have included
medical mistrust due to systemic medical
racism\textsuperscript{\protect\hyperlink{ref-stoler2021}{17},\protect\hyperlink{ref-mosby2021}{22}},
mistrust in
vaccines\textsuperscript{\protect\hyperlink{ref-willis2021}{15}}, and
the existence of health misinformation and
disinformation\textsuperscript{\protect\hyperlink{ref-mosby2021}{22}--\protect\hyperlink{ref-freeman2020}{24}}.
However, it is important also to consider that vaccination uptake can be
influenced by geographical (spatial) factors. In this regard,
differences in COVID-19 vaccination rates have been associated with
varied regional attitudes towards
vaccination\textsuperscript{\protect\hyperlink{ref-malik2020}{25}}, lack
of vaccine access in areas where underrepresented groups
reside\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7}}, as well
as an absence of prioritization of areas inhabited by vulnerable
groups\textsuperscript{\protect\hyperlink{ref-nguyen2021}{26}}. Other
studies have also shown heterogeneity in vaccine uptake within small
governmental administrative units such as
counties\textsuperscript{\protect\hyperlink{ref-mollalo2021}{27}--\protect\hyperlink{ref-bhuiyan2022}{30}},
indicating that accounting for geographical differences in vaccination
can help predict patterns of booster
uptake\textsuperscript{\protect\hyperlink{ref-wood2022}{31}}. However,
there is a limited amount of studies that have analyzed geographical
differences in vaccination in Canada. Existing studies in this area have
focused on differences within certain cities, such as
Toronto\textsuperscript{\protect\hyperlink{ref-choi2021}{32}}, or
Montréal\textsuperscript{\protect\hyperlink{ref-mckinnon2021}{33}}), or
have explored differences between
provinces\textsuperscript{\protect\hyperlink{ref-guay2022}{19}}, but to
our knowledge, there are no studies that analyze differences in
vaccination uptake at the provincial level. Such need is specially
important in the context of Canada's pandemic response goals, which have
been to minimize serious illness and deaths while minimizing societal
disruption\textsuperscript{\protect\hyperlink{ref-naci2023}{34}}.
Analyzing differences in vaccination uptake within the provinces can aid
to identify inequalities that might exist and that need to be addressed
before the advent of another pandemic.

This need is especially important in the case of Ontario, the most
populated province of Canada, and which as has a complex healthcare
system. Between 2007 and 2019, Ontario managed healthcare access to its
inhabitants using 14 intra-provincial divisions called the Local Health
Integration Networks (LHINs), which aimed to provide an integrated
health system. However, this approach was complex, bureaucratic,
resulting in excessive expenditures, disparities in mortality rates, the
deterioration of certain performance indicators (such as wait times and
hospital readmissions), fragmented electronic health systems, and
inequities in health services
access\textsuperscript{\protect\hyperlink{ref-tsasis2012}{35}--\protect\hyperlink{ref-lysyk2016}{39}}.
With the intent of better organizing and delivering care, in late 2019
the provincial government eliminated the LHINs and incorporated the
areas covered by them into six larger Health Regions (North East, North
West, Central, Toronto, West, and East), which are managed by a new
government agency, Ontario Health
(OH)\textsuperscript{\protect\hyperlink{ref-dong2022}{37}}.

On the other hand, public health in Ontario is administered by Public
Health Ontario (PHO), a government agency established in late 2007 and
that is currently composed of 34 Public Health Units (PHUs) that cover
the entire geography of the province. Although PHUS were commissioned
with leading the distribution of the COVID-19 vaccine within their
respective
areas\textsuperscript{\protect\hyperlink{ref-rotenberg2021}{40}}, the
vaccine rollout occurred with significant interaction between PHO and
OH; in many instances both agencies had to work together to organize
vaccination clinics, with personnel associated with OH also being
actively involved when demand exceeded the capacity of PHO personnel,
especially in rural
areas\textsuperscript{\protect\hyperlink{ref-ashcroft2023}{41}}. Indeed,
based on the experience of COVID-19, there is an ongoing debate on how
OH and PHO will interact in the future, and the challenges that this
will entail for the healthcare system of the
province\textsuperscript{\protect\hyperlink{ref-dion2022}{42}}.

Therefore, considering the relatively recent adoption of the Health
Region model and its alignment with the onset of the COVID-19 pandemic,
there is an ongoing need to analyze the existence of geographical
disparities in vaccination uptake within the Health Regions and identify
the socio-demographic groups that might be affected, as this can serve
as an indirect assessment of the state of the implementation of the new
model while helping identify ongoing challenges that decision-makers
might need to address to ensure the long-term success of this model and
its interaction with PHO, which is specially important considering that
previous research has highlighted disparities in the level of activity
of each Health
Region\textsuperscript{\protect\hyperlink{ref-sethuram2023}{43}}.

Therefore, in this study we hypothesized that there were differences in
COVID-19 vaccination rates between the Health Regions between October
2021 and January of 2022. To understand and contextualize these
differences, we by included socio-economic factors in our analysis,
aiming to identify in which demographic groups were particularly
impacted, in order to provide an assessment of the current state of
healthcare access in Ontario.

@@

We used data from the \emph{Survey of COVID-19 related Behaviours and
Attitudes}, a repeated cross sectional survey focused on the Canadian
province of Ontario that was commissioned by the Fields Institute for
Research in Mathematical Sciences and the Mathematical Modelling of
COVID-19 Task Force under ethical guidance from the University of
Toronto (under protocol 00043317), and which ran between September 30th,
2021 and January 17th,2022.

Briefly, the survey was deployed using random domain intercept
technology, a methodology where internet users whose device meta-data
indicated their presence in the province of Ontario had a random chance
of being redirected to the survey after they had clicked on a registered
but commercially inactive web link, or typed in a web address for a site
that was dormant but that was temporarily managed by the company that
administered the survey (RIWI
Corp)\textsuperscript{\protect\hyperlink{ref-sargent2022}{44}}. Users
then decided whether to anonymously participate in the survey, and those
that participated were able to exit the survey at any time. After the
survey closed, regardless if it was complete or incomplete, access was
denied to any further users with the same internet protocol address
(IP), effectively allowing each user only one opportunity to participate
in the survey. Users who indicated they were under the age of 16 were
exited from the survey without creating a record. Finally, the personal
identifier information from each user that participated in the survey
was automatically scrubbed and replaced by a unique ID.

The survey allowed users to enter their socio-economic information , and
asked information on vaccination status by using the question ``Have you
received the first dose of the COVID vaccine?'', with possible answers
``yes'' and ``no'' (Table~\ref{tbl-descriptive-stats}). Additionally,
the survey automatically collected the geographical location of the
respondent (using the nearest municipality, as shown in
Figure~\ref{fig-map}), and the date of access to the survey. The
original dataset contained 39,029 observations with a response rate of
16.25\% (defined as observations that had complete answers), due to the
survey design that allowed respondents to exit at any time and deployed
the questions randomly. We selected 6,343 observations that were
labelled as ``complete'' in the dataset and that corresponded to those
that had answers for all covariates considered in our analysis. Later,
we matched the city of each observation with its corresponding LHIN and
Health Region, and removed observations from areas with low
representation (254 observations corresponding to the North West and
North East Health Regions). After all the preliminary analyses, the
total number of observations used for analysis was 6,236 and included
the East, Central, Toronto, and West Health Regions covering between
October 1st,2021 and January 17, 2022. The original dataset, clean
dataset, and details on the data cleaning process are described in
detail in the \href{https://github.com/aimundo/Fields_COVID-19/}{GitHub
repository} for this paper.

@@

We used a logistic regression model to examine the impact of the Health
Regions in vaccination rates while considering the socio-economic
factors and months covered by the survey
(Table~\ref{tbl-descriptive-stats}) and certain interactions (Race and
Health Region and Race and income), as previous studies have shown that
socio-economic factors and their interactions are significant predictors
of intent of vaccination and vaccination
status\textsuperscript{\protect\hyperlink{ref-nguyen2022}{45}--\protect\hyperlink{ref-cnat2022a}{47}}.
Because we identified differences in representativity between the survey
data and the estimates from the Census, we used an iterative
proportional fitting procedure
(\emph{raking})\textsuperscript{\protect\hyperlink{ref-deming1940}{48}}
to correct the data using data from the Census and Health Region
population totals; and fitted the regression model to the uncorrected
and corrected data. Details regarding the correction can be found in the
Appendix. All analyses were conducted in R 4.2.2 using the packages
\texttt{survey}\textsuperscript{\protect\hyperlink{ref-lumley2011}{49}},\texttt{tidyverse}\textsuperscript{\protect\hyperlink{ref-wickham2019}{50}},
\texttt{quarto}\textsuperscript{\protect\hyperlink{ref-quarto}{51}},
\texttt{modelsummary}\textsuperscript{\protect\hyperlink{ref-modelsummary}{52}},
and
\texttt{gtsummary}\textsuperscript{\protect\hyperlink{ref-gtsummary}{53}}.

@@

Our results show that indeed, there were differences in vaccination odds
across Ontario in certain socio-demographic groups. Specifically, those
who identified as Arab/Middle Eastern, Black, Latin American, having
mixed racial or ethnic background, or that belonged to other groups not
explicitly included in the survey (Southeast Asian, Filipino, West
Asian, and minority groups not identified elsewhere) had vaccination
odds that were between a third and a half of that of individuals that
identified as White or Caucasian (Figure~\ref{fig-models}). These
results are consistent with previous studies that have shown lower
vaccination rates in individuals with the same socio-demographic
characteristics\textsuperscript{\protect\hyperlink{ref-guay2022}{19}--\protect\hyperlink{ref-hussain2022}{21},\protect\hyperlink{ref-carter2022}{54}}.

Lower vaccine uptake in the socio-demographic groups indicated above may
be influenced in part, by vaccine hesitancy and refusal, which have been
associated in underrepresented Canadian individuals with concerns on
vaccine safety, effectiveness, and experiences of racial discrimination
in health
settings\textsuperscript{\protect\hyperlink{ref-cnat2022a}{47},\protect\hyperlink{ref-basta2022}{55}--\protect\hyperlink{ref-cnat2023}{57}}.
However, it has been shown that structural barriers also play an
important role in vaccination uptake. In the case of underrepresented
individuals, such barriers include complex scheduling systems, language
barriers, lack of adequate public transportation, and lack of accessible
vaccination
sites\textsuperscript{\protect\hyperlink{ref-njoku2021}{58}}. In this
regard, it is interesting to note that vaccination venues were scarce in
low socio-economic areas that had the highest burden of COVID-19 in
Toronto and other regions of Ontario around the time covered by the
survey\textsuperscript{\protect\hyperlink{ref-bogoch2022}{7},\protect\hyperlink{ref-iveniuk2021}{59}},
and that pharmacies in the Peel region (an area identified as a
``hotspot'' with high numbers of essential workers and multigenerational
households) could not keep up with vaccine
demand\textsuperscript{\protect\hyperlink{ref-gill2022}{60}}. This
suggests that the observed differences are associated with disparities
in vaccine access that were present during the period covered by the
survey.

@@

This result is likely reflects in part the fact that individuals in
underrepresented groups tend to perform occupations that have been
deemed as ``essential'' in the context of the
pandemic\textsuperscript{\protect\hyperlink{ref-hawkins2020}{61},\protect\hyperlink{ref-ct2021}{62}},
which include workers in the areas of grocery stores, gas stations,
warehouses, distribution, and manufacturing, all being occupations for
which an income within the significant brackets identified in the
analysis is to be expected. In Ontario, these workers had priority for
COVID-19
vaccination\textsuperscript{\protect\hyperlink{ref-mishra2021}{63}}; and
there is evidence of interventions by vaccination staff in certain parts
of the province to encourage vaccination uptake by these
individuals\textsuperscript{\protect\hyperlink{ref-gill2022}{60}}. These
facts, combined with evidence of increased trends in vaccination in this
group
elsewhere\textsuperscript{\protect\hyperlink{ref-nguyen2021b}{64}},
suggest that the type of occupation of individuals in underrepresented
groups played an important role in increasing the odds of vaccination.

@@

First, in this case it is useful to analyze the data considering the
LHINs in each Health Region, because most studies in the literature
focused on Ontario use the LHINs as the base of their analyses. The West
Health Region covers the area previously occupied by the Hamilton
Niagara Haldimand Brant, South West, and Waterloo Wellington LHINs,
whereas the East Health Region covers the area of the former Champlain
and Central East LHINs. Previous research has identified health
disparities in these (mostly rural) regions, such as unequal
distribution of primary care providers, increased mortality, and low
pharmacist
availability\textsuperscript{\protect\hyperlink{ref-shah2019}{65}--\protect\hyperlink{ref-timony2022}{67}}.

Furthermore, there is an ongoing challenge for the health system of the
province with regard to personalized healthcare for marginalized
individuals. For example, the West Health Region has only two Aboriginal
Health Access Centres (community-led primary healthcare organizations
focused on First Nations, Métis, and Inuit communities) to provide care
to an estimated 100,000 Indigenous individuals living in the
area\textsuperscript{\protect\hyperlink{ref-ontariohealth}{68}}. Lack of
access to personalized healthcare affects individuals that may mistrust
the traditional healthcare system due to systemic racism or oppression,
which is known to be the case for Indigenous and Black individuals in
Canada, as these rationales have been associated with observed lower
vaccination rates among these
groups\textsuperscript{\protect\hyperlink{ref-smylie2022}{69},\protect\hyperlink{ref-eissa2021}{70}}.
Taken together, this suggests that healthcare disparities specific to
these underrepresented groups in certain parts of the province impacted
vaccination uptake, and highlights the need of investments in the Health
Regions focused on resources, infrastructure, and specially personnel
that can deliver personalized care to marginalized communities, as it
has been shown that such efforts have improved trust in vaccination in
underrepresented groups
elsewhere\textsuperscript{\protect\hyperlink{ref-schafferderoo2020}{71}}.

@@

Additionally, our analysis did not consider the North West and North
East Health Regions, due to the low number of entries from these areas
in the survey (Figure~\ref{fig-map}). Low representation is expected as
these regions as they only account for 5\% of the total population of
Ontario. However, these areas have the highest proportion of Indigenous
inhabitants\textsuperscript{\protect\hyperlink{ref-ontariohealth}{68}}.
In the context of personalized care, there is a need to collect data
that focuses on these Health Regions where additional health disparities
might be present and possibly understudied.

The results in this study are based on self-reported data, where bias
might be present. However, in the context of COVID-19, it has been shown
that good agreement exists between self-reported and documented
vaccination
status\textsuperscript{\protect\hyperlink{ref-stephenson2022}{72}}, we
believe that our data was able to provide a valid sample of vaccination
uptake in the province. This is supported by the statistically
significant higher vaccination odds that were identified for January of
2022 in the model, which are consistent with province-wide trends
reported by Public Health Ontario (which show a 4\% increase between
early December and January, in contrast to a 2.5\% increase between
October and
November\textsuperscript{\protect\hyperlink{ref-ontario-covid}{73}});
however, the short time window constitutes essentially a ``snapshot''
view of the evolution of the disease, and additional data would be
needed to obtain estimates per racial/ethnic group over time across all
Health Regions that can help inform the existence of other health
disparities.

@@

\leavevmode\vadjust pre{\hypertarget{ref-naci2023}{}}%
\CSLLeftMargin{34. }%
\CSLRightInline{National Advisory Committee on Immunization (NACI).
\emph{Guidance on an Additional COVID-19 Booster Dose in the Spring of
2023 for Individuals at High Risk of Severe Illness Due to COVID-19}.
Public Health Agency of Canada; 2023.
\url{https://www.canada.ca/en/public-health/services/publications/vaccines-immunization/national-advisory-committee-immunization-guidance-additional-covid-19-booster-dose-spring-2023-individuals-high-risk-severe-illness-due-covid-19.html}}

\leavevmode\vadjust pre{\hypertarget{ref-tsasis2012}{}}%
\CSLLeftMargin{35. }%
\CSLRightInline{Tsasis P, Evans JM, Owen S. Reframing the challenges to
integrated care: A complex-adaptive systems perspective.
\emph{International Journal of Integrated Care}. 2012;12(5).
doi:\href{https://doi.org/10.5334/ijic.843}{10.5334/ijic.843}}

\leavevmode\vadjust pre{\hypertarget{ref-muratov2018}{}}%
\CSLLeftMargin{36. }%
\CSLRightInline{Muratov S, Lee J, Holbrook A, et al. Regional variation
in healthcare spending and mortality among senior high-cost healthcare
users in ontario, canada: A retrospective matched cohort study.
\emph{{BMC} Geriatrics}. 2018;18(1).
doi:\href{https://doi.org/10.1186/s12877-018-0952-7}{10.1186/s12877-018-0952-7}}

\leavevmode\vadjust pre{\hypertarget{ref-dong2022}{}}%
\CSLLeftMargin{37. }%
\CSLRightInline{Dong L, Sahu R, Black R. Governance in the
transformational journey toward integrated healthcare: The case of
ontario. \emph{Journal of Information Technology Teaching Cases}.
Published online December 2022:204388692211473.
doi:\href{https://doi.org/10.1177/20438869221147313}{10.1177/20438869221147313}}

\leavevmode\vadjust pre{\hypertarget{ref-lysyk2015}{}}%
\CSLLeftMargin{38. }%
\CSLRightInline{Auditor General of Ontario O of the, ed. Annual {R}eport
2015. In: \emph{Section 3.08: LHINs - Local Health Integration
Networks}. Queen's Printer for Ontario; 2015. Accessed May 12, 2023.
\url{https://www.auditor.on.ca/en/content/annualreports/arreports/en15/3.08en15.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-lysyk2016}{}}%
\CSLLeftMargin{39. }%
\CSLRightInline{Auditor General of Ontario O of the, ed. Annual {R}eport
2016. In: \emph{Section 3.03: Electronic Health Records' Implementation
Status}. Queen's Printer for Ontario; 2016. Accessed May 12, 2023.
\url{https://www.auditor.on.ca/en/content/annualreports/arreports/en16/v1_303en16.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-rotenberg2021}{}}%
\CSLLeftMargin{40. }%
\CSLRightInline{Rotenberg S, Downer MB, Brown H, et al. \emph{{COVID}-19
Vaccination for People with Disabilities}. Ontario {COVID}-19 Science
Advisory Table; 2021.
doi:\href{https://doi.org/10.47326/ocsat.2021.02.35.1.0}{10.47326/ocsat.2021.02.35.1.0}}
